---
title: "Agentic Workflows & GitHub"
subtitle: "EMSE 4572/6572: Exploratory Data Analysis"
format: html
toc: true
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| warning: false

knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  fig.path = "figs/",
  fig.width = 7.252,
  fig.height = 4,
  comment = "#>",
  fig.retina = 3
)

# If you don't have these packages installed, run these lines to do so:
# install.packages('tidyverse')
# install.packages('cowplot')

# Load libraries
library(tidyverse)
library(cowplot)
```

# Practice 1: GitHub Desktop

## Checkpoint: your solo loop from HW1

Confirm that:

- Your `eda-practice` repo shows up in **GitHub Desktop**
- The same repo exists on **GitHub.com**

If any of these fail, flag it now. The steps are in [HW1](https://eda.seas.gwu.edu/2026-Fall/hw/1-agentic-workflows-and-github.html).

## Commit to your partner's repo

Pair up with a neighbor and work through the collaboration loop. You'll both push straight to the `main`branch.

1. In *your* `eda-practice` repo on GitHub.com: **Settings › Collaborators** → add your partner's GitHub username → they accept the emailed invite
2. Swap repo links. In GitHub Desktop: **File › Clone Repository** → clone *your partner's* repo
3. Edit their `README.md` in Positron → **Commit** to `main` → **Push**
4. Back in *your own* repo, click **Fetch origin**, then **Pull**
5. Open `README.md` — your partner's edit should be there

Write down the URL of your partner's repo, and what you changed in it:

> Your answer here

Look at your repo's **History** in GitHub Desktop. How can you tell which
commit was your partner's, and what it changed?

> Your answer here

# Practice 2: Your first agentic project

Direct an agent to build a multi-page Quarto website about **someone you
admire** — a scientist, athlete, artist, whoever.

1. In GitHub Desktop, make a new repo called `my-website` (include a `README`
   file). Make it **public**, and publish it to GitHub.
2. In Positron: **File › Open Folder…** and pick that repo's folder.
3. In the terminal pane, type `claude` and press Enter.
4. Prompt Claude for a multi-page **Quarto website** about your person.
5. When it's done, ask it to render the site (if it didn't already).
6. Open `index.html` in the `_site` folder.
7. **Commit + push** to GitHub.

## Extra: publish with GitHub Pages

1. Ask Claude to change the Quarto output folder from `_site` to `docs` (edit
   `output-dir` in `_quarto.yml`), then re-render.
2. **Commit + push**.
3. On GitHub.com: **Settings › Pages** → **Source: Deploy from a branch**,
   **Branch: `main`**, **Folder: `/docs`** → **Save**.
4. Wait a minute, then visit the URL GitHub shows at the top of that page.

Your site's URL:

> Your answer here

# Practice 3: Modes & slash commands

Still in your `my-website` repo:

1. Run `/init` — then **open the `CLAUDE.md` it wrote and read it.**
2. Run `/model` and `/effort` to see your options.
3. Ask for a **small** change (a new page, a navbar tweak) at `/effort low`.
4. Press **shift + tab** into **plan mode**, then ask for something harder — e.g. *"collapse everything into a single-page website"* — at `/effort high`. Read the plan before you approve it.
5. `/compact`, then keep going. Does it still know what you're building?
6. `/clear`, then ask *"what am I working on?"*
7. **Commit + push**.

# Practice 4: Skills

A **skill** is a slash command you define that teaches the agent one of your recipes. Today's class folder ships one: `my-chart-style`.

1. Ask Claude to copy the `my-chart-style` folder from this class folder into `.claude/skills/` in your project (create the folder if it doesn't exist).
2. Come up with a chart you'd like to see from `data/flights.csv`.
3. Ask for that chart **without** the skill, then again invoking **`/my-chart-style`**.
4. Modify the skill — a specific color for all charts, a different theme, whatever — and re-create the chart.

The data:

```{r}
flights <- read_csv(file.path('data', 'flights.csv'))
airlines <- read_csv(file.path('data', 'airlines.csv'))
```

Have the agent create a plot **without** the skill here:

```{r}
# Write code here
```

Then have the agent create a plot **with** the `/my-chart-style` skill here:

```{r}
# Write code here
```

# Practice 5: Which airline should you avoid?

The code will run, the chart will probably look great. **But the answer may be wrong.**

Ask Claude to rank the airlines by **mean departure delay** — joining
`flights.csv` to `airlines.csv` for the names — and chart it.

```{r}
#| label: airline-delay-ranking

delay_by_airline <- flights |>
  left_join(airlines, by = "carrier") |>
  group_by(carrier, name) |>
  summarize(
    mean_dep_delay = mean(dep_delay, na.rm = TRUE),
    .groups = "drop"
  ) |>
  arrange(desc(mean_dep_delay))

delay_by_airline

delay_by_airline |>
  ggplot(aes(x = mean_dep_delay, y = reorder(name, mean_dep_delay))) +
  geom_col(fill = "#80C5DCFF", width = 0.8) +
  scale_x_continuous(expand = expansion(mult = c(0, 0.05))) +
  labs(
    x = "Mean departure delay (minutes)",
    y = NULL,
    title = "Mean departure delay varies across airlines",
    subtitle = "Cancelled flights are excluded from the mean",
    caption = "Data source: nycflights13 flights and airlines tables"
  ) +
  theme_minimal_vgrid(font_family = "sans") +
  theme(
    plot.title.position = "plot",
    plot.caption.position = "plot",
    plot.caption = element_text(hjust = 0, face = "italic"),
    panel.background = element_rect(fill = "white", color = NA),
    plot.background = element_rect(fill = "white", color = NA)
  )
```

Which airline does the chart say is worst?

> Your answer here

## Now interrogate it

Don't ask the agent whether it's right Check it a way it can't fake.

**How many flights is each bar calculated from?**

```{r}
# Write code here
```

> Your answer here

**What happened to the cancelled flights?** (A cancelled flight has `NA` for `dep_delay`.) How many did each airline have, and what share of its flights is that?

```{r}
# Write code here
```

> Your answer here

**Is any airline missing from the chart?** Compare the carriers in `flights` to the carriers in `airlines`, and to the ones that actually made it into the plot.

```{r}
# Write code here
```

> Your answer here

## Fix it

Tell Claude *specifically* what's wrong, and have it fix the chart.

```{r}
#| fig-height: 5
#| fig-width: 7

# Write code here
```

Does the fixed chart change your answer about which airline to avoid?

> Your answer here

# Verification checklist

Run this every time — it's the skill this whole course is really about.

- [ ] Did **row / column counts** change the way I expected?
- [ ] Did I **read the code**, not just the summary?
- [ ] Did I **spot-check** a few values against the source file?
- [ ] Were rows **silently dropped** (`NA` handling, inner vs. left join)?
- [ ] Do **totals / aggregates** make sense?
- [ ] Is it **reproducible**, or did it hard-code an answer?
